\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2017-01-10 pea_paper_docs/tex/pea.tex
added egl to tissue conditional analysis
and wrote the paragraph for it…..

\subsection*{Deconstructing the \emph{egl} phenotype}
How does a phenotype emerge? We realized that with the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes associated
with a particular phenotype, then search for tissue terms that are enriched to
understand how a phenotype arises from interactions between anatomical regions.
As a test of this, we selected the \emph{egg-laying defective} (\emph{egl}) phenotype.
\emph{A priori}, it is not clear what tissues may contribute the most towards
an \emph{egl} phenotype. In \cel{}, egg-laying is a complex behavior that involves
a large number of tissues. The somatic gonad acts as a repository for the eggs,
the uterine seam cells help protect the uterus, and a variety of muscles help
contract the uterus and open the vulva to lay an egg. The vulva must be well-formed
to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) helps the
animal sense lay eggs, possibly by sensing pressure on the uterus. The
complexity of the interactions that happen to allow egg-laying make understanding
the \emph{egl} phenotype in terms of tissues a challenging task.

We extracted all of the \cel{} genes that have been associated with an \emph{egl}
phenotype and we used TEA to understand what tissues are enriched. The `HSN'
was the most enriched tissue contributing to \emph{egl} phenotypes (\qval{13}).
Next, the vulD and vulC vulval cells were most enriched respectively (\qval{7}).
The somatic gonad was fourth most enriched (\qval{6}). The next biggest contribution
comes from the Pn.p cells and the vulE, vulF cells (\qval{6}), followed by the
vulA, vulB2 and vulB1 cells (\qval{5}). The uterine muscles are enriched at lower
levels (\qval{2}), comparably to the uterine seam cells. Therefore, the \emph{egl}
phenotype would seem to emerge primarily from defects in sensing (via the HSN),
secondarily from defects in the vulva, and only sometimes from defects in the
uterine seam cells or muscles. It is notable that the vulva was not equally
enriched. Although all the `vul' cells are annotated to a similar degree (between
50--70 genes for each cell type), the vulD and vulE cells were considerably more
likely to be associated with an \emph{egl} phenotype than the others. This may
reflect the fact that vulD and vulE are the site of attachment for four vulval
muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
to these cells prevents adequate function of these muscles. However, vulF is
also attached to a set of four additional vulval muscles, vm2. Why
is vulF less associated with an \emph{egl} phenotype?

\section*{Conclusions}

>>>>

\subsection*{Deconstructing the Egl phenotype}
How does a phenotype emerge? We realized that with the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes associated
with a particular phenotype, then search for tissue terms that are enriched to
understand how a phenotype arises from interactions between anatomical regions.
As a test of this, we selected the egg-laying defective (Egl) phenotype.
In \cel{}, egg-laying is a complex behavior that involves
a large number of tissues~\cite{}. The somatic gonad acts as a repository for the eggs,
the uterine seam cells help protect the uterus, and a variety of muscles help
contract the uterus and open the vulva to lay an egg~\cite{}. The vulva must be well-formed
to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
in the egg-laying control~\cite{}. The complexity of the interactions that
happen to allow egg-laying make understanding the Egl phenotype in terms of
tissues a challenging task.

We extracted all of the \cel{} genes that have been associated with an Egl
phenotype and we used TEA to understand what tissues are enriched. The HSN
was enriched more than five-fold above background (\qval{7}) as were vulD, vulC,
vulE and vulF (\qval{6}). The vulA, vulB2 and vulB1 were
enriched at slightly lower levels (\qval{5}), whereas the uterine muscles and
uterine seam cells were enriched more than twice above background levels
(\qval{2}).
Therefore, the Egl phenotype would seem to emerge primarily from defects in the
HSN, secondarily from defects in the vulva, and only sometimes from defects in the
uterine seam cells or muscles. It is notable that all vulval cells were not equally
enriched. Although all the `vul' cells are annotated to a similar degree (between
50--70 genes for each cell type), the vulD and vulC cells had the largest
enrichment effect size and the lowest q-values, suggesting that these cells
are more likely to be associated with an Egl phenotype than the others. This may
reflect the fact that vulD and vulE are the site of attachment for four vulval
muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
to these cells prevent adequate function of these muscles. In support of these
observations, P7.pa had the largest fold-enrichment of any tissue. In \cel{},
P7.pa gives rise to vulD and vulC. However, vulF is
also attached to a set of four additional vulval muscles, vm2. Why
is vulF less associated with an Egl phenotype?

\section*{Conclusions}
The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
an important step towards a unified set of analyses that can help researchers
to understand genomic datasets. By offering tissue, phenotype and gene ontology
enrichment on a single site, researchers will benefit both in the speed at which
the analysis is completed, but also from the fact that the analyses are all
carried out with the same methodology and the same underlying algorithms.
